---
title: "Taxi Project Descriptive Analysis"
author: "Brad Dixon <rbdixon@gmail.com>"
date: "May 7, 2015"
output: html_document
---

```{r init, include=FALSE}
library(ProjectTemplate)
load.project()
library(knitr)
opts_chunk$set(warning=FALSE, fig.width=10, fig.height=10)
```

# Data Sets

## Train

```{r train}
print(glimpse(TRAIN_RAW))
```

## Test

```{r test}
print(glimpse(TEST_RAW))
```

# Variables

## TRIP_ID

> TRIP_ID: (String) It contains an unique identifier for each trip;

Assumptions:

* Trip IDs are assigned to each row in train and test data sets.
* Trip IDs are unique across train and test data sets.

```{r TRIP_ID}
DATA %>% 
  group_by(dataset) %>% 
  filter(is.na(TRIP_ID)) %>% 
  summarize(n = n()) %>% 
  filter(n > 1) %>% 
  group_by(dataset) %>% 
  summarize(num_trip_id_NA = n()) %>% 
  collect

DATA %>%
  group_by(dataset, TRIP_ID) %>% 
  summarize(n = n()) %>% 
  filter(n > 1) %>% 
  group_by(dataset) %>% 
  summarize(num_trip_id_duplicated = n()) %>% 
  collect
```

Finding:

* TEST, TRAIN: No NA TRIP_ID's.
* TRAIN: Trip IDs are duplicated.

## CALL_TYPE

> CALL_TYPE: (char) It identifies the way used to demand this service. It may contain one of three possible values:
> ‘A’ if this trip was dispatched from the central;
> ‘B’ if this trip was demanded directly to a taxi driver on a specific stand;
> ‘C’ otherwise (i.e. a trip demanded on a random street).

Assumptions:

* No missing values
* A, B, and C levels are found in both train and test
* No unexpected values

```{r CALL_TYPE}
DATA %>% 
  group_by(dataset, CALL_TYPE) %>% 
  summarize(n = n()) %>% 
  arrange(dataset, CALL_TYPE) %>% 
  collect %>% 
  spread(CALL_TYPE, n)
```

Finding:

* All assumptions confirmed.

## ORIGIN_CALL

> ORIGIN_CALL: (integer) It contains an unique identifier for each phone number which was used to demand, at least, one service. It identifies the trip’s customer if CALL_TYPE=’A’. Otherwise, it assumes a NULL value;

Assumptions:

* Most "A" type taxi calls have a ORIGIN_CALL identifier assigned.
* ORIGIN_CALL identifiers are persistent to the phone as indicated by some identifiers taking repeated cab rides.
* Some ORIGIN_CALL identifiers will be found in training and test sets.

```{r ORIGIN_CALL}
# How many records have ORIGIN_CALL assigned by CALL_TYPE?
DATA %>% 
  mutate(
    origin_call_assigned = ORIGIN_CALL!="" & !is.na(ORIGIN_CALL)
  ) %>% 
  group_by(dataset, CALL_TYPE, origin_call_assigned) %>% 
  summarize(n = n()) %>% 
  collect %>% 
  spread(origin_call_assigned, n)

# How many distinct ORIGIN_CALL ID's have taken multiple trips?
DATA %>% 
  group_by(dataset, ORIGIN_CALL) %>% 
  summarize(multiple_trips = n()>1) %>% 
  group_by(dataset, multiple_trips) %>% 
  summarize(n = n()) %>% 
  arrange(dataset, multiple_trips) %>% 
  collect %>% 
  spread(multiple_trips, n)

# How many ORIGIN_CALL ID's are found in both data sets?
DATA %>%
  filter(!is.na(ORIGIN_CALL), ORIGIN_CALL!="") %>% 
  select(dataset, ORIGIN_CALL) %>% 
  distinct %>% 
  group_by(ORIGIN_CALL) %>% 
  summarize(
    in_train_and_test = any(dataset=="TRAIN") & any(dataset=="TEST")
  ) %>% 
  group_by(in_train_and_test) %>% 
  summarize(n=n()) %>% 
  collect

# Top ORIGIN_CALL ID's
DATA %>% 
  filter(!is.na(ORIGIN_CALL), ORIGIN_CALL!="") %>% 
  group_by(dataset, ORIGIN_CALL) %>% 
  summarize(n=n()) %>% 
  filter(n>1) %>% 
  group_by(dataset) %>% 
  top_n(10, n) %>% 
  collect
```

Findings:

* TEST, TRAIN: Empty string or NA indicates missing ORIGIN_CALL assignment.
* TEST, TRAIN: Only CALL_TYPE=="A" have ORIGIN_CALL assigned.
* The train data set has many repeat trips by identified phone numbers (ORIGIN_CALL).
* The test data set is mostly one-trip per phone number. This makes sense since test is multiple snapshots in time.
* Very few ORIGIN_CALL ID's are found in both train and test.
* Some ORIGIN_CALL ID's are very frequent callers?
* ORIGIN_CALL identifier 2002 is a very frequent taxi requester. Possibly a hotel or a veiled NA?

## ORIGIN_STAND

> ORIGIN_STAND: (integer): It contains an unique identifier for the taxi stand. It identifies the starting point of the trip if CALL_TYPE=’B’. Otherwise, it assumes a NULL value

Assumptions:

* Most ORIGIN_STAND == "B" calls will have a taxi stand identifier but not all will.
* Stands are frequent repeat callers of taxis.

```{r ORIGIN_STAND}
# How frequently is ORIGIN_STAND assigned by CALL_TYPE?
DATA %>% 
  group_by(dataset, CALL_TYPE) %>% 
  mutate(
    origin_stand_assigned = ORIGIN_STAND!="" & !is.na(ORIGIN_STAND)
  ) %>% 
  group_by(dataset, CALL_TYPE, origin_stand_assigned) %>% 
  summarize(n=n()) %>% 
  collect %>% 
  arrange(dataset, CALL_TYPE) %>% 
  spread(origin_stand_assigned, n)

# Top ORIGIN_STANDs
DATA %>% 
  group_by(dataset, ORIGIN_STAND) %>% 
  summarize(n=n()) %>% 
  filter(n>1) %>% 
  top_n(10, n) %>% 
  collect %>% 
  print.data.frame()
```

Findings:

* TRAIN, TEST: ORIGIN_STAND can be either NA or "".
* TEST: There are examples of CALL_TYPE=="B" and ORIGIN_STAND in train but none are found in TEST.

## TAXI_ID

> TAXI_ID: (integer): It contains an unique identifier for the taxi driver that performed each trip;

Assumptions:

* TAXI_ID will not be missing

```{r TAXI_ID}
# Are TAXI_ID's always assigned?
DATA %>% 
  group_by(dataset) %>% 
  mutate(
    taxi_id_assigned = !is.na(TAXI_ID)
  ) %>% 
  group_by(dataset, taxi_id_assigned) %>% 
  summarize(n=n()) %>% 
  collect %>% 
  spread(taxi_id_assigned, n)

# Top Taxi ID's
DATA %>% 
  group_by(dataset, TAXI_ID) %>% 
  summarize(n=n()) %>% 
  filter(n>1) %>% 
  top_n(10, n) %>% 
  collect %>% 
  print.data.frame()
```

Findings:

* Assumptions confirmed.
* TEST: Most taxi drivers are observed once in the test data.

## TIMESTAMP

> TIMESTAMP: (integer) Julian Timestamp (in seconds). It identifies the trip’s start

Assumptions:

* Origin is 1970-01-01. This means it isn't really a Julian timestamp.
* Timezone is UTC

Testing data set date range:

> Five test sets will be available to evaluate your predictive framework (in one single CSV file named "test.csv"). Each one of these datasets refer to trips that occurred between 01/07/2014 and 31/12/2014. Each one of these data sets will provide a snapshot of the current network status on a given timestamp. It will provide partial trajectories for each one of the on-going trips during that specific moment.
> The five snapshots included on the test set refer to the following timestamps:
> 
> * 14/08/2014 18:00:00
> * 30/09/2014 08:30:00
> * 06/10/2014 17:45:00
> * 01/11/2014 04:00:00
> * 21/12/2014 14:30:00

```{r TIMESTAMP_TT}
DATA %>% 
  group_by(dataset) %>% 
  summarize(
    mindt = to_timestamp(min(TIMESTAMP)),
    maxdt = to_timestamp(max(TIMESTAMP))
  ) %>% 
  collect
```

```{r TIMESTAMP_TEST}
TEST_RAW %>% 
  mutate(
    dt = to_timestamp(TIMESTAMP)
  ) %>% 
  collect %>% 
  ggplot(aes(dt)) +
    geom_histogram() +
    xlab("Date") +
    ylab("Count of Trip Records") +
    ggtitle("TEST Dataset Trips by Day") +
    theme_bw()

TEST_RAW %>% 
  mutate(
    dt_day = date_trunc('day', to_timestamp(TIMESTAMP))
  ) %>%
  group_by(dt_day) %>% 
  arrange(dt_day) %>% 
  summarize(
    mindt = to_timestamp(min(TIMESTAMP)),
    maxdt = to_timestamp(max(TIMESTAMP)) 
  ) %>% 
  collect
```

Findings:

* TEST: Time skewing? This does not appear to be UTC since the 6/10/2014 data, gathered at 17:45:00, has a maxdate timestamp of 13:44:33. Discussions in the forum indicate otherwise. Odd. The first 4 days are skewed by 4 hours and the last by 5 hours.

## DAY_TYPE

> DAYTYPE: (char) It identifies the daytype of the trip’s start. It assumes one of three possible values:
>
> * ‘B’ if this trip started on a holiday or any other special day (i.e. extending holidays, floating holidays, etc.);
> * ‘C’ if the trip started on a day before a type-B day;
> * ‘A’ otherwise (i.e. a normal day, workday or weekend).

Assumptions:

* All levels are represented in both data sets

```{r DAY_TYPE}
DATA %>% 
  group_by(dataset, DAY_TYPE) %>% 
  summarize(n=n()) %>% 
  collect %>% 
  spread(DAY_TYPE, n)
```

Findings:

* Only "A" days (normal) days are represented.

## MISSING_DATA

> MISSING_DATA: (Boolean) It is FALSE when the GPS data stream is complete and TRUE whenever one (or more) locations are missing

Assumption:

* Some days are missing from both data sets

```{r MISSING_DATA}
DATA %>% 
  group_by(dataset, MISSING_DATA) %>% 
  summarize(n=n()) %>% 
  collect %>% 
  spread(MISSING_DATA, n)
```

Finding:

* Only the training data has trips with missing data and only 10 trips are missing data all together.

## POLYLINE

> POLYLINE: (String): It contains a list of GPS coordinates (i.e. WGS84 format) mapped as a string. The beginning and the end of the string are identified with brackets (i.e. [ and ], respectively). Each pair of coordinates is also identified by the same brackets as [LONGITUDE, LATITUDE]. This list contains one pair of coordinates for each 15 seconds of trip. The last list item corresponds to the trip’s destination while the first one represents its start;

POLYLINE is mapped into a trajectory table (`TRAJ`) by PostGIS SQL statements. The field `GH` is a 9 character geohash of each point in the trajectory.

```{r polyline_gh4}
TRAJ
```