---
title: "Map Analysis"
author: "Brad Dixon <rbdixon@gmail.com>"
date: "May 12, 2015"
output:
  html_document:
    keep_md: yes
---

```{r init, include=FALSE}
library(ProjectTemplate)
load.project()
library(knitr)
library(leaflet)
opts_chunk$set(warning=FALSE, fig.width=10, fig.height=10)
```

# Porto, Portugal

```{r porto}
map=leaflet() %>% 
  addTiles(options = tileOptions(opacity=0.4)) %>% 
  setView(-8.621953, 41.162142, 12)
map
```

# Trip Variables

```{r trip_scatterplots}
FULL_TRIP_DETAILS = TRIP_DETAILS %>% 
  # filter(level==9) %>% 
  left_join(GHINFO, by=c(origin_gh="gh")) %>% 
  rename(level=level.x, origin_centroid=centroid, origin_lat=lat, origin_lon=lon, origin_geo=geo, origin_geojson=geojson) %>% 
  select(-level.y, -gh) %>% 
  left_join(GHINFO, by=c(dest_gh="gh")) %>% 
  select(-level.y, -gh) %>% 
  rename(level=level.x, dest_centroid=centroid, dest_lat=lat, dest_lon=lon, dest_geo=geo, dest_geojson=geojson) %>% 
  compute

FULL_TRIP_DETAILS %>% 
  filter(level==9) %>% 
  select(az, dist, recs, origin_lat, origin_lon, dest_lat, dest_lon) %>% 
  collect %>% 
  ggpairs()
```

# Lat/Lon Scatterplot

```{r trip_map}
ORIGIN=FULL_TRIP_DETAILS %>%
  filter(level==9) %>% 
  group_by(origin_gh) %>% 
  summarize(
    n = n(),
    lat=mean(origin_lat),
    lon=mean(origin_lon),
    dist = mean(dist)
  ) %>% 
  collect

DESTINATION=FULL_TRIP_DETAILS %>%
  filter(level==9) %>% 
  group_by(dest_gh) %>% 
  summarize(
    n = n(),
    lat=mean(dest_lat),
    lon=mean(dest_lon),
    dist = mean(dist)
  ) %>% 
  collect

map %>% 
  addCircles(ORIGIN$lon, ORIGIN$lat, color="green") %>% 
  addCircles(DESTINATION$lon, DESTINATION$lat, color="blue")
```

# Origin / Destination Pairs

```{r od_curve}
ODPAIRS = FULL_TRIP_DETAILS %>% 
  filter(level %in% 4:9) %>% 
  group_by(level, origin_gh, dest_gh) %>% 
  summarize(n=n()) %>% 
  group_by(level) %>% 
  arrange(desc(n)) %>% 
  mutate(
    pct = n/sum(n),
    cumpct = cumsum(n)/sum(n)
  ) %>% 
  mutate(
    num = row_number()
  ) %>% 
  collect

ODPAIRS %>% 
  ggplot(aes(num, cumpct)) +
    geom_line() +
    facet_wrap(~ level) +
    theme_bw()
```

Above is CDF of trips by geohash. Not much of a hook past level 7.

```{r od_heatmap}
ODPAIRS %>% 
  filter(level==4) %>% 
  arrange(desc(pct)) %>% 
  ggplot(aes(origin_gh, dest_gh, fill=pct)) +
    geom_raster() +
    theme_bw()
```

Above shows heatmap of level 4 origin/destination pairs.

```{r od_heatmap_ez3f}
ODPAIRS %>% 
  filter( 
    substring(origin_gh, 1, 4) == "ez3f",
    level==5
  ) %>% 
  arrange(desc(pct)) %>% 
  ggplot(aes(origin_gh, dest_gh, fill=pct)) +
    geom_raster() +
    theme_bw()
```

Zoom in on ez3f.

```{r od_heatmap_ez3f_f5_fh}
ODPAIRS %>% 
  filter( 
    substring(origin_gh, 1, 5) %in% c("ez3f5", "ez3fh"),
    level==6
  ) %>% 
  arrange(desc(pct)) %>% 
  ggplot(aes(origin_gh, dest_gh, fill=pct)) +
    geom_raster() +
    theme_bw()
```

Zoom in on e3f5 and e3fh.

# Self-routing

Origin and Destination geohashes are the same.

```{r od_by_level_selfroute_cumpct}
SELFROUTE = ODPAIRS %>% 
  group_by(level) %>% 
  mutate(
    total_n = sum(n)
  ) %>% 
  filter(origin_gh==dest_gh) %>% 
  arrange(desc(n)) %>% 
  mutate(
    cumpct = cumsum(n)/total_n,
    num = row_number()
  )

SELFROUTE %>% 
  group_by(level) %>% 
  summarise_each(funs(min, mean, max), pct)

ggplot(SELFROUTE, aes(num, cumpct)) +
  geom_line() +
  facet_wrap(~ level) +
  theme_bw()
```

~25% of trips self-route on level 5. Past that there isn't much self-routing.

```{r which_selfroute_level4}
SRL5 = SELFROUTE %>% 
  filter(level==5, pct>.1) %>% 
  left_join(GHINFO, by=c(origin_gh="gh", level="level"), copy=TRUE) %>% 
  collect %>% 
  print()

map %>% 
  addGeoJSON(geojsoncol_to_fc(SRL5$geojson))
```

The above two level 5 gehashes account for ~26% of trips.